Reparametrization of COM-Poisson Regression Models with Applications in the Analysis of Experimental Data

### 2018-09-13 article.Rnw
Make the changes suggested by John Hinde

\Affiliations{
\item Department of Exact Sciences,
  University of S\~{a}o Paulo - ESALQ,
  Piracicaba,
  SP, Brazil
\item Department of Statistics,
  Paran\'{a} Federal University,
  Curitiba,
  PR, Brazil
\item School of Mathematics, Statistics and Applied Mathematics,
  National University of Ireland,
  Galway,
  Galway, Ireland
}

@@

%-------------------------------------------
% Title and shor title
\Title{Reparametrization of COM-Poisson Regression Models with
  Applications in the Analysis of Experimental Data}
\TitleRunning{Reparametrization of COM-Poisson Regression Models}

% -------------------------------------------
% Abstract
\Abstract{The COM-Poisson distribution is a two-parameter generalization
  of Poisson distribution can deal with under-, equi- and overdispersed
  count data. Its location parameter does not correspond to the
  expectation, which complicates the interpretation of regression models
  specified using this distribution. In this paper, we propose a
  straightforward reparametrization of the COM-Poisson distribution
  based on an approximation to the expectation of this distribution. The
  main advantage of our new parametrization is the interpretation of the
  regression coefficients in terms of the expectation of the count
  response variable, as usual in the context of generalized linear
  models. Furthermore, the estimation and inference for the
  reparametrized COM-Poisson regression models can be done based on the
  likelihood paradigm. We carried out simulation studies to verify the
  finite sample properties of the maximum likelihood estimators. The
  results from our simulation study show that the maximum likelihood
  estimators are unbiased and consistent for both regression and
  dispersion parameters. The empirical correlation between the
  regression and dispersion parameter estimators and deviance surfaces
  suggests that these parameters are orthogonal. We illustrate the
  application of the proposed model through the analysis of three data
  sets with over-, under- and equidispersed count data. The study of
  distribution properties through a consideration of dispersion,
  zero-inflated and heavy tail indexes, together with the results of
  data analysis show the flexibility over standard approaches.  The
  computational routines for fitting the original and new version of the
  COM-Poisson regression model and the analyzed data sets are available
  in the supplementary material.  }

@@

Count data are random variables that assume non-negative integer values
and represent the number of times an event occurs in the observation
period. This kind of data is common in crop sciences, such as the number
of grains produced by a plant, number of fruits produced by a tree,
number of insects captured by a trap, to cite but a few.  Since the
seminal paper of \citet{Nelder1972} where the class of the generalized
linear models (GLMs) was introduced, the analysis of count data often
employs the Poisson regression model. This model provides a suitable
strategy for the analysis of count data and an efficient Newton scoring
algorithm can be used for fitting the model.

In spite of the advantages of the Poisson regression model, the Poisson
distribution has only one parameter, which represents both the
expectation and variance of the count random variable.  This restriction
on the relationship between the expectation and variance induced by the
Poisson distribution is referred as equidispersion.  However, in
practical data analysis such restriction can be unsuitable, since the
observed data can present  variance both smaller or larger than the
expectation, leading to the cases of under- and overdispersion,
respectively.  The main problem of the application of the Poisson
regression model to non-equidispersed count data is that the standard
errors associated with the regression coefficients are inconsistently
estimated, which in turn can lead to misleading
inferences~\citep{Winkelmann1995, Bonat2017}.

In practice, overdispersion is largely reported in the literature and
may occur due to the absence of relevant covariates, heterogeneity of
sampling units, different observational periods/regions not considered
in the analysis, and excess of zeros~\citep{Hinde1998}.  The case of
underdispersion is less report in the literature, however, it has been
of increasing interest in the statistical community. The processes that
reduce the variability are not as well-known as those leading to extra
variability.  For this reason, there are few approaches to deal with
underdispersed count data. The explanatory mechanisms leading to
underdispersion may be related to the underlying stochastic process
generating the count data.  When the time between events is not
exponentially distributed, the number of events can be over or
underdispersed, this process motivated the class of duration dependence
models~\citep{Winkelmann1995}.  Another possible explanation of
underdispersion is when the responses correspond to order statistics of
component observations, such as maxima of Poisson distributed
counts~\citep{Steutel1989}.

The strategies for constructing alternative count distributions are
related with the causes of the non-equidispersion. Specifically for the
overdispersion, Poisson mixture models are widely applied.  One
popular example of this approach is the negative-binomial model, where
the expectation of the Poisson distribution is assumed to be gamma
distributed.  However, other distributions can be used to represent the
random variation.  For example the Poisson-Tweedie
model~\citep{Bonat2017} and its special cases as the Poisson
inverse-Gaussian and Neyman-Type A assume that the random effects are
Tweedie, inverse Gaussian and Poisson distributed, respectively.  The
Gamma-Count distribution assumes a gamma distribution for the time
between events, thus it can deal with underdispersed as well as
overdispersed count data~\citep{Zeviani2014}. The COM-Poisson
distribution is obtained by a generalization of the Poisson distribution
allowing for a non-linear decrease in the ratios of successive
probabilities~\citep{Shmueli2005}.

The COM-Poisson distribution has the Poisson and geometric distributions
as special cases, as well as the Bernoulli distribution as a limiting
case. It can deal with both under- and overdispersed count data.  Some
recently applications of the COM-Poisson distribution include
\citet{Lord2010} for the analysis of traffic crash data,
\citet{Sellers2010} for the modelling of airfreight breakage and book
purchases, \citet{Huang2017} to the analysis of attendance data,
takeover birds and cotton boll counts, and \citet{Chatla2018} to model
counts from bike sharing systems. Theoretical results and approximations
derived from this distribution is discussed by \citet{Shmueli2005},
\citet{Daly2016} and \citet{Gaunt2017}. The main disadvantage of the
COM-Poisson regression model as presented in~\citet{Sellers2010} is that
its location parameter does not correspond to the expectation of the
distribution, which complicates the interpretation of regression models
and means that they are not comparable with standard approaches such as
the Poisson and negative binomial regression models. \citet{Huang2017}
proposed a mean-parametrization of the COM-Poisson distribution in order
to avoid such an issue.  In this approach the mean parameter is obtained
by solving an non-linear equation defined as an infinite sum.
Consequently, it is computationally demanding and liable to numerical
problems.

The main goal of this article is to propose a novel COM-Poisson
parametrization based on the mean approximation presented by
\citet{Shmueli2005}. In this parametrization, the probability mass
function is written in terms of $\mu$ and $\phi$, where $\mu$ is the
approximated expectation and $\phi$ is a dispersion parameter. In
contrast to the original parametrization, the proposed parametrization
leads to regression coefficients directly associated with the
expectation of the response variable, as usual in the context of
generalized linear models.  Consequently, the obtained regression
coefficients are comparable with the ones obtained by standard
approaches, such as the Poisson and negative binomial regression models.
Furthermore, our novel COM-Poisson parametrization is simpler than the
strategy proposed by~\cite{Huang2017}, since it does not require any
numerical method for solving non-linear equations, and we show the
attractive properties like the orthogonality between dispersion and
regression parameters and consistency and asymptotic normality of the
maximum likelihood estimators are retained.

This paper is organized as follows. In \autoref{background} we present
the COM-Poisson distribution and the strategy proposed
by~\cite{Huang2017}.  The proposed reparametrization, assessment of
moment approximations, and study of distribution properties are
considered in the~\autoref{reparametrization}. In
the~\autoref{estimation-and-inference} we present estimation and
inference for the novel COM-Poisson regression model based on the
likelihood paradigm. The properties of the maximum likelihood estimators
and the orthogonality property are assessed
in~\autoref{simulation-study} through simulation studies.  We illustrate
the application of the new COM-Poisson regression model through the
analysis of three data sets. We provide an \texttt{R} implementation of
the COM-Poisson and reparameterized COM-Poisson regression models as
well as the analyzed data sets in the supplementary
material.\footnote{Available on
  \texttt{\url{http://www.leg.ufpr.br/~eduardojr/papercompanions}}
  \label{papercompanion}.}.

@@

The COM-Poisson distribution generalizes the Poisson distribution in
terms of the ratio between the probabilities of two consecutive events
by adding an extra dispersion parameter~\citep{Sellers2010}.  Let $Y$ be
a COM-Poisson random variable, then
$$\frac{\Pr(Y=y-1)}{\Pr(Y=y)} = \frac{y^\nu}{\lambda}$$ while for the
Poisson distribution this ratio is $\frac{y}{\lambda}$ corresponding to
$\nu=1$.  It allows the COM-Poisson distribution deals with
non-equidispersed count data.  The probability mass function of the
COM-Poisson distribution is given by
\begin{equation}
  \label{eqn:pmf-cmp}
  \Pr(Y=y \mid \lambda, \nu) = \frac{\lambda^y}{(y!)^\nu Z(\lambda, \nu)},
  \qquad y = 0, 1, 2, \ldots,
\end{equation}
where $\lambda > 0$, $\nu \geq 0$ and $Z(\lambda, \nu) =
\sum_{j=0}^\infty \frac{\lambda^j}{(j!)^\nu}$ is a normalizing
constant that depends on both parameters.

The $Z(\lambda, \nu)$ series diverges theoretically only when $\nu=0$
and $\lambda \geq 1$, but numerically for small values of $\nu$ combined
with large values of $\lambda$, the sum is so huge it causes
overflow. \autoref{tab:convergenceZ} shows the values of the normalizing
constants using one thousand increments, that is,
$\sum_{j=0}^{1000}\lambda^j/(j!)^\nu$ for different values of $\lambda$
and $\phi$.

@@

\begin{table}[ht]
\centering
\caption{Values for $Z(\lambda, \nu)$ constant (numerically computed) for values of $\lambda$ (0.5 to 50) and $\phi$ (0 to 1)}
\label{tab:convergenceZ}
\begingroup\small
\begin{tabularx}{\textwidth}{C|CCCCCC}
  \toprule

@@

\footnotesize \raggedright
divergent$^{*}$ is a mathematically divergent series; and
divergent$^{**}$ a numerically divergent series.
\end{table}

In the first line of \autoref{tab:convergenceZ} we have mathematically
divergent series, because $\sum_{j=0}^\infty\lambda^j$ is divergent when
$\lambda \geq 1$.  In other cases the series diverges numerically, due
to the computational storage limitation.  For both forms of divergence
it is impossible to compute probabilities, therefore, this acts as a
restriction on the parameter space.

An undesirable feature of the COM-Poisson distribution is that the
moments cannot be obtained in closed form. \citet{Shmueli2005} and
\citet{Sellers2010} using an asymptotic approximation for
$Z(\lambda,\nu)$, showed that the expectation and variance of the
COM-Poisson distribution can be approximated by
\begin{equation}
  \label{eqn:mean-approx}
  \text{E}(Y) \approx \lambda^{1/\nu} - \frac{\nu - 1}{2\nu} \qquad
  \textrm{and} \qquad
  \textrm{Var}(Y) \approx \frac{\lambda^{1/\nu}}{\nu}\,,
\end{equation} which is particularly accurate for  $\nu \leq 1$ or
$\lambda > 10$. The authors also argue that the mean-variance
relationship can be approximate by
$\frac{1}{\nu}\text{E}(Y)$. In~\autoref{reparametrization}, we assess
the accuracy of these approximations.

The COM-Poisson regression model was proposed by~\citet{Sellers2010},
using the original parametrization. In this case, the COM-Poisson
regression model is $\log(\lambda_i) = \bm{x}_i^\top \bm{\beta}$ and the
relationship between E$(Y_i)$ and $\bm{x}_i$ is modelled indirectly.
\citet{Huang2017} shows how to model directly the expectation of the
COM-Poisson distribution in a suitable parametrization. In the
\autoref{eqn:pmf-cmp}, \Citeauthor{Huang2017} proposes that the
parameter $\lambda$ as a function of $\mu$ and $\nu$, is given by the
solution to
\begin{equation*}
  \sum_{j=0}^{\infty} (j - \mu) \frac{\lambda^j}{(j!)^\nu} = 0\,.
\end{equation*}
Thus the mean-parametrized COM-Poisson regression model is $\log(\mu_i)
= \bm{x}_i^\top \bm{\beta}$. In this article, we propose an alternative
mean-parametrization of the COM-Poisson distribution in order to avoid the
limitations of the original parametrization and the numerical complexity
of the \Citeauthor{Huang2017}'s approach.

@@

The proposed reparametrization of COM-Poisson models is based on
the mean approximation (\autoref{eqn:mean-approx}). We introduced a new
parameter $\mu$, using this approximation,
\begin{equation}
  \label{eqn:repar-cmp}
  \mu = h_\nu(\lambda) = \lambda^{1/\nu} - \frac{\nu - 1}{2\nu}
  \quad \Rightarrow \quad
  \lambda = h_\nu^{-1}(\mu) = \left (\mu +
    \frac{(\nu - 1)}{2\nu} \right )^\nu.
\end{equation}
The dispersion parameter is taken on the log scale for computational
convenience, thus $\phi = \log(\nu)$, $\phi \in \mathbb{R}$. The
interpretation of $\phi$ is the same as the $\nu$, but on another
scale. For $\phi < 0$ and $\phi > 0$ we have the overdispersed and
underdispersed cases, respectively.  When $\phi=0$ we have Poisson
distribution as a special case.

@@

In order to assess the accuracy of the moment approximations
(\autoref{eqn:mean-approx}), \autoref{fig:approx-plot} presents the
quadratic errors for (a) expectation and (b) variance.  The quadratic
errors were obtained by $[\mu - \text{E}(Y)]^2$ for the expectation and
by $[ \mu \exp(-\phi) - \text{Var}(Y)]^2$ for the variance. In both
cases $\text{E}(Y)$ and Var$(Y)$ were computed numerically\footnote{We
  compute E$(Y) = \sum_{y=0}^{500}yp(y)$ and
  E$(Y^2) = \sum_{y=0}^{500}y^2p(y)$.}. The dotted lines represent the
border between the regions $\nu \leq 1$ and $\lambda > 10^\nu$, in the
$\mu$ and $\phi$ scale.

<<approx-plot, fig.height=3, fig.width=6, out.width=".8\\textwidth", fig.cap="Quadratic errors for the approximation of the (a) expectation and (b) variance. Dotted lines represent the restriction for suitable approximations given by \\cite{Shmueli2005}.">>=
@

The results in \autoref{fig:approx-plot} show that the mean
approximation is accurate, the largest quadratic error is
\Sexpr{max(grid[["emu"]])} for the parameter values evaluated. For the
variance approximation, the largest quadratic error was
\Sexpr{max(grid[["eva"]])} and it occurs for negative values of
$\phi$. Interestingly, the errors are larger for negative values of
$\phi$ and present no clear relation with $\mu$, as opposed to the
regions gives by \citet{Shmueli2005} ($\phi \leq 0$ and
$\mu > 10 - \frac{\exp(\phi) - 1}{2\exp(\phi)}$).

The results presented in \autoref{fig:approx-plot}(a) support the
proposed reparametrization. Replacing $\lambda$ and $\nu$ as function of
$\mu$ and $\phi$ in \autoref{eqn:pmf-cmp}, the reparametrized
distribution takes the form
\begin{equation}
  \label{eqn:pmf-cmpmu}
  \Pr(Y=y \mid \mu, \phi) =
  \left ( \mu +\frac{ e^\phi-1}{2e^\phi} \right )^{ye^\phi}
  \frac{(y!)^{-e^\phi}}{Z(\mu, \phi)},
  \qquad y = 0, 1, 2, \ldots\,,
\end{equation}
where $\mu > 0$. We denote this distribution as
COM-Poisson$_\mu$. In \autoref{fig:pmf-cmp}, we show the shapes of
COM-Poisson$_\mu$ distribution.

@@

The constraint $\mu>0$ imposes an undesirable restriction in the
original parameter space, $\lambda^\nu>(\nu-1)/(2\nu)$. However, this
restricted region is related to small underdispersed counts (averages
smaller than 0.1 and $\nu>1$) a parameter region unexplored in practice.

In order to explore the flexibility of the COM-Poisson model to deal
with real count data, we compute indexes for dispersion (DI),
zero-inflation (ZI) and heavy-tail (HI), which are respectively given by
\begin{equation*}
\text{DI} = \frac{\text{Var}(Y)}{\text{E}(Y)}, \quad
\text{ZI} = 1 + \frac{\log \Pr(Y = 0)}{\text{E}(Y)}
  \quad \text{and} \quad
\text{HT} = \frac{\Pr(Y=y+1)}{\Pr(Y=y)}\quad \text{for} \quad y \to
\infty.
\end{equation*}
These indexes are defined in relation to the Poisson distribution. Thus,
the dispersion index indicates overdispersion for $\text{DI} > 1$,
underdispersion for $\text{DI} < 1$ and equidispersion for
$\text{DI} = 1$. The zero-inflation index indicates zero-inflation for
$\text{ZI} > 0$, zero-deflation for $\text{ZI} < 0$ and no excess of
zeros for $\text{ZI} = 0$. Finally, heavy-tail index indicates a
heavy-tail distribution for $\text{HT} \to 1$ when $y \to \infty$.

These indexes are discussed by \citet{Bonat2017} to study the
flexibility of Poisson-Tweedie distribution, and \citet{Puig2006} to
describe count distributions. Regarding the COM-Poisson$_\mu$
distribution, in \autoref{fig:indexes-plot} we present the relationship
between (a) mean and variance, (b--c) the dispersion and zero-inflation
indexes for different values of $\mu$ and $\phi$, and (d) heavy-tail
index for $\mu=25$ and different values of $y$ and $\phi$.

@@

<<indexes-plot, fig.width=9.5, fig.height=4, fig.cap="Indexes for COM-Poisson distribution. (a) Mean and variance relationship, (b--d) dispersion, zero-inflation and heavy-tail indexes for different parameter values. Dotted lines represents the Poisson special case.">>=
@

>>>>

\Affiliations{
\item Department of Exact Sciences,
  University of S\~{a}o Paulo - ESALQ,
  Piracicaba,
  SP, Brazil
\item Department of Statistics,
  Paran\'{a} Federal University,
  Curitiba,
  PR, Brazil
\item School of Mathematics, Statistics and Applied Mathematics,
  National University of Ireland Galway,
  Galway, Ireland
}

@@

%-------------------------------------------
% Title and short title
\Title{Reparametrization of COM-Poisson Regression Models with
  Applications in the Analysis of Experimental Data}
\TitleRunning{Reparametrization of COM-Poisson Regression Models}

% -------------------------------------------
% Abstract
\Abstract{The COM-Poisson distribution is a two-parameter generalization
  of the Poisson distribution that can deal with under-, equi- and
  overdispersed count data. Unfortunately, its location parameter does
  not correspond to the expectation, which complicates the
  interpretation of regression models specified using this
  distribution. In this paper, we propose a straightforward
  reparametrization of the COM-Poisson distribution based on an
  approximation to the expectation. The main advantage of our new
  parametrization is the interpretation of the regression coefficients
  in terms of the (approximate) expectation of the count response
  variable, as is usual in the context of generalized linear
  models. Estimation and inference for the reparametrized COM-Poisson
  regression models can be done using the likelihood paradigm. We
  carried out simulation studies to verify the finite sample properties
  of the maximum likelihood estimators. The results from our simulation
  study show that the maximum likelihood estimators are unbiased and
  consistent for both regression and dispersion parameters. The
  empirical correlation between the regression and dispersion parameter
  estimators and nature of the deviance surfaces suggests that these
  parameters are also approximately orthogonal, which is advantageous
  for interpretation, inference, and computational efficiency. We
  illustrate the application of the proposed model through the analysis
  of three data sets with over-, under- and equidispersed count
  data. The study of the distribution's properties, through a
  consideration of dispersion, zero-inflation and heavy tail indexes,
  together with the results of data analysis, show the flexibility over
  standard approaches.  The computational routines for fitting the
  original and reparameterized versions of the COM-Poisson regression
  model and data sets are available in the supplementary material.  }

@@

Count data arise from random variables that take non-negative integer
values and typically represent the number of times an event occurs in an
observation period or region. This kind of data is also common in crop
sciences, examples including the number of grains produced by a plant,
the number of fruits on a tree, the number of insects captured by a
trap, etc.  Since the seminal paper of \citet{Nelder1972}, where the
class of the generalized linear models (GLMs) was introduced, the
analysis of count data often uses a Poisson regression model. This model
provides a suitable strategy for the analysis of count data and an
efficient Newton scoring algorithm that can be used for fitting.

In spite of the advantages of the Poisson regression model, the Poisson
distribution has only one parameter that represents both the expectation
and variance of the count random variable.  This equality restriction on
the relationship between the expectation and variance of the Poisson
distribution is referred as equidispersion.  However, in practical data
analysis such restriction can be unsuitable, since the observed data can
present variance both smaller or larger than the mean, leading to under-
and overdispersion, respectively.  While applying the Poisson regression
model to non-equidispersed count data can give consistent parameter
estimates, the associated standard errors are inconsistently estimated,
which in turn can lead to misleading inferences~\citep{Winkelmann1995,
  Bonat2017}.

In practice, overdispersion is widely reported in the literature and may
occur due to the absence of relevant covariates, heterogeneity of
sampling units, different observational periods/regions not being
considered in the analysis, and excess zeros~\citep{Hinde1998}.  The
case of underdispersion is less often reported in the literature,
however, it has been of increasing interest in the statistical
community. The processes that reduce the variability are not as
well-known as those leading to extra variability.  For this reason,
there are few models to deal with underdispersed count data. Possible
explanatory mechanisms leading to underdispersion may be related to the
underlying stochastic process generating the count data.  For example,
when the time between events is not exponentially distributed, the
number of events can be over or underdispersed; a process that motivated
the class of duration dependence models~\citep{Winkelmann1995}.  Another
possible explanation for underdispersion is when the responses
correspond to order statistics of component observations, such as maxima
of Poisson distributed counts~\citep{Steutel1989}.

Strategies for constructing alternative count distributions are related
to the causes of the non-equidispersion. Specifically for
overdispersion, Poisson mixture (compound) models are widely applied.
One popular example of this approach is the negative-binomial model,
where the expectation of the Poisson distribution is assumed to be gamma
distributed.  However, other distributions can also be used to represent
this additional random variation.  For example the Poisson-Tweedie
model~\citep{Bonat2017} and its special cases the Poisson
inverse-Gaussian and Neyman-Type A models assume that the random effects
are Tweedie, inverse Gaussian or Poisson distributed, respectively.  The
Gamma-Count distribution assumes a gamma distribution for the time
between events and it can deal with underdispersed as well as
overdispersed count data~\citep{Zeviani2014}. The subject of this paper,
the COM-Poisson distribution, is obtained by a generalization of the
Poisson distribution that allows for a non-linear decrease in the ratios
of successive probabilities~\citep{Shmueli2005}.

The COM-Poisson distribution includes the Poisson and geometric
distributions as special cases, as well as the Bernoulli distribution as
a limiting case. It can deal with both under- and overdispersed count
data.  Some recent applications of the COM-Poisson distribution include
\citet{Lord2010} for the analysis of traffic crash data,
\citet{Sellers2010} for the modelling of airfreight breakage and book
purchases, \citet{Huang2017} on the analysis of attendance data,
takeover bids and cotton boll counts, and \citet{Chatla2018} to model
counts from bike sharing systems. Theoretical results and approximations
derived for this distribution are discussed by \citet{Shmueli2005},
\citet{Daly2016} and \citet{Gaunt2017}. The main disadvantage of the
COM-Poisson regression model as presented in~\citet{Sellers2010} is that
its location parameter does not correspond to the expectation of the
distribution. This complicates the interpretation of regression models
and means that they are not comparable with standard approaches, such as
the Poisson and negative binomial regression models. In order to avoid
this issue, \citet{Huang2017} proposed a mean-parametrization of the
COM-Poisson distribution.  In his approach the mean parameter is
obtained as the solution of a non-linear equation defined as an infinite
sum, which is computationally demanding and liable to numerical
problems.

The main goal of this article is to propose a novel COM-Poisson
parametrization based on the mean approximation presented by
\citet{Shmueli2005}. In this parametrization, the probability mass
function is written in terms of $\mu$ and $\phi$, where $\mu$ is the
approximate expectation and $\phi$ is a dispersion parameter. In
contrast to the original parametrization, the proposed parametrization
leads to regression coefficients directly associated (approximately)
with the expectation of the response variable, as is usual in the
context of generalized linear models.  Consequently, the resulting
regression coefficients are comparable to those obtained by standard
approaches, such as the Poisson and negative binomial regression models.
Furthermore, our novel COM-Poisson parametrization is computationally
simpler than the strategy proposed by~\citet{Huang2017}, since it does
not require any numerical method for solving non-linear equations. Also,
we show that attractive properties like the orthogonality between
dispersion and regression parameters and consistency and asymptotic
normality of the maximum likelihood estimators are retained.

This paper is organized as follows. In \autoref{background} we present
the COM-Poisson distribution and the approach proposed
by~\citet{Huang2017}.  The newly proposed reparametrization is
considered in ~\autoref{reparametrization}, along with assessment of the
moment approximation, and a study of distribution's
properties. In~\autoref{estimation-and-inference} we present estimation
and inference for the novel COM-Poisson regression model in a likelihood
framework. The properties of the maximum likelihood estimators and their
approximate orthogonality are assessed in~\autoref{simulation-study}
through simulation studies.  We illustrate the application of the new
COM-Poisson regression model with the analysis of three data sets. We
provide an \texttt{R} implementation for the COM-Poisson and
reparameterized COM-Poisson regression models, together with the
analyzed data sets, in the supplementary material.\footnote{Available on
  \texttt{\url{http://www.leg.ufpr.br/~eduardojr/papercompanions}}
  \label{papercompanion}.}.

@@

The COM-Poisson distribution generalizes the Poisson distribution in
terms of the ratio between the probabilities of two consecutive events
by adding an extra dispersion parameter~\citep{Sellers2010}.  Let $Y$ be
a COM-Poisson random variable, then
$$\frac{\Pr(Y=y-1)}{\Pr(Y=y)} = \frac{y^\nu}{\lambda}$$ while for the
Poisson distribution this ratio is $\frac{y}{\lambda}$ corresponding to
$\nu=1$.  This allows the COM-Poisson distribution to deal with
non-equidispersed count data.  The probability mass function of the
COM-Poisson distribution is given by
\begin{equation}
  \label{eqn:pmf-cmp}
  \Pr(Y=y \mid \lambda, \nu) = \frac{\lambda^y}{(y!)^\nu Z(\lambda, \nu)},
  \qquad y = 0, 1, 2, \ldots,
\end{equation}
where $\lambda > 0$, $\nu \geq 0$ and
$Z(\lambda, \nu) = \sum_{j=0}^\infty \frac{\lambda^j}{(j!)^\nu}$ is a
normalizing constant that depends on both parameters.

The $Z(\lambda, \nu)$ series diverges theoretically only when $\nu=0$
and $\lambda \geq 1$, but numerically, for small values of $\nu$
combined with large values of $\lambda$, the sum is so huge it results
in overflow. \autoref{tab:convergenceZ} shows the values of the
normalizing constant based on one thousand terms in the summation, that
is $\sum_{j=0}^{1000}\lambda^j/(j!)^\nu$, for different values of
$\lambda$ and $\phi$.

@@

\begin{table}[ht]
\centering
\caption{Values for $Z(\lambda, \nu)$ normalizing constant (computed numerically with 1000 terms in the summation) for values of $\lambda$ (0.5 to 50) and $\phi$ (0 to 1)}
\label{tab:convergenceZ}
\begingroup\small
\begin{tabularx}{\textwidth}{C|CCCCCC}
  \toprule

@@

\footnotesize \raggedright
divergent$^{*}$ denotes a mathematically divergent series; and
divergent$^{**}$ a numerically divergent series.
\end{table}

In the first line of \autoref{tab:convergenceZ} we have mathematically
divergent series, because $\sum_{j=0}^\infty\lambda^j$ is divergent when
$\lambda \geq 1$.  In other cases the series diverges numerically, due
to the computational storage limitation.  For both forms of divergence
it is impossible to compute COM-Poisson probabilities, therefore, this
places a restriction on the feasible parameter space.

An undesirable feature of the COM-Poisson distribution is that the
moments cannot be obtained in closed form. \citet{Shmueli2005} and
\citet{Sellers2010} using an asymptotic approximation for
$Z(\lambda,\nu)$, showed that the expectation and variance of the
COM-Poisson distribution can be approximated by
\begin{equation}
  \label{eqn:mean-approx}
  \text{E}(Y) \approx \lambda^{1/\nu} - \frac{\nu - 1}{2\nu} \qquad
  \textrm{and} \qquad
  \textrm{Var}(Y) \approx \frac{\lambda^{1/\nu}}{\nu}\,,
\end{equation}with greatest accuracy for  $\nu \leq 1$ or
$\lambda > 10$. The authors also argue that the mean-variance
relationship can be approximated as
$\textrm{Var}(Y) \approx\frac{1}{\nu}\text{E}(Y)$.
In~\autoref{reparametrization}, we assess the accuracy of these
approximations.

The COM-Poisson regression model was proposed by~\citet{Sellers2010},
using this original parametrization. Specifically, the COM-Poisson
regression model is defined as
$\log(\lambda_i) = \bm{x}_i^\top \bm{\beta}$ and the relationship
between E$(Y_i)$ and $\bm{x}_i$ is modelled indirectly.
\citet{Huang2017} shows how to model directly the expectation of the
COM-Poisson distribution in a suitable reparametrization. From
\autoref{eqn:pmf-cmp}, \Citeauthor{Huang2017} notes that the parameter
$\lambda$ is given, as a function of $\mu$ and $\nu$, by the solution to
\begin{equation*}
  \sum_{j=0}^{\infty} (j - \mu) \frac{\lambda^j}{(j!)^\nu} = 0\,.
\end{equation*}
This allows \Citeauthor{Huang2017} to define a mean-parametrized
COM-Poisson regression model with
$\log(\mu_i) = \bm{x}_i^\top \bm{\beta}$ giving a direct relationship
between $\mu$ and the covariates $\bm{x}$. In this article, we propose
an alternative approximate mean-parametrization of the COM-Poisson
distribution in order to avoid the limitations of the original
parametrization and the numerical complexity of the
\Citeauthor{Huang2017}'s approach.

@@

The proposed reparametrization of COM-Poisson models is based on the
mean approximation (\autoref{eqn:mean-approx}). We introduce a new
parameter $\mu$, using this approximation,
\begin{equation}
  \label{eqn:repar-cmp}
  \mu = h_\nu(\lambda) = \lambda^{1/\nu} - \frac{\nu - 1}{2\nu}
  \quad \Rightarrow \quad
  \lambda = h_\nu^{-1}(\mu) = \left (\mu +
    \frac{(\nu - 1)}{2\nu} \right )^\nu.
\end{equation}
The dispersion parameter is taken on the log scale for computational
convenience, thus $\phi = \log(\nu)$, $\phi \in \mathbb{R}$. The
interpretation of $\phi$ is the same as the $\nu$, but simply on another
scale. For $\phi < 0$ and $\phi > 0$ we have the overdispersed and
underdispersed cases, respectively.  When $\phi=0$ we have the Poisson
distribution as a special case.

@@

In order to assess the accuracy of the moment approximations
(\autoref{eqn:mean-approx}), \autoref{fig:approx-plot} presents the
squared errors for (a) expectation and (b) variance.  The squared errors
are $[\mu - \text{E}(Y)]^2$ for the expectation and
$[ \mu \exp(-\phi) - \text{Var}(Y)]^2$ for the variance. In both cases
$\text{E}(Y)$ and Var$(Y)$ were computed numerically using 500
terms\footnote{ E$(Y) \approx \sum_{y=0}^{500}yp(y)$ and
  E$(Y^2) \approx \sum_{y=0}^{500}y^2p(y)$.}. The dotted lines represent
the border between the regions $\nu \leq 1$ and $\lambda > 10^\nu$, in
the $\mu$ and $\phi$ scale.

<<approx-plot, fig.height=3, fig.width=6, out.width=".8\\textwidth", fig.cap="Squared errors for the approximation of the (a) expectation and (b) variance. Dotted lines represent the restriction for suitable approximations given by \\cite{Shmueli2005}.">>=
@

The results in \autoref{fig:approx-plot} show that the mean
approximation is accurate, the largest squared error, for the parameter
values evaluated, is $0.038$.  For the variance approximation, the
largest squared error was $XXX$ and it occurs for negative values of
$\phi$. Interestingly, the errors for the variance are larger for
negative values of $\phi$ and present no clear relation with $\mu$, as
opposed to the regions gives by \citet{Shmueli2005} ($\phi \leq 0$ and
$\mu > 10 - \frac{\exp(\phi) - 1}{2\exp(\phi)}$).

The results presented in \autoref{fig:approx-plot}(a) support the
proposed reparametrization. Replacing $\lambda$ and $\nu$ as functions of
$\mu$ and $\phi$ in \autoref{eqn:pmf-cmp}, the reparametrized
distribution takes the form
\begin{equation}
  \label{eqn:pmf-cmpmu}
  \Pr(Y=y \mid \mu, \phi) =
  \left ( \mu +\frac{ e^\phi-1}{2e^\phi} \right )^{ye^\phi}
  \frac{(y!)^{-e^\phi}}{Z(\mu, \phi)},
  \qquad y = 0, 1, 2, \ldots\,,
\end{equation}
where $\mu > 0$. We denote this reparameterized distribution as
COM-Poisson$_\mu$. In \autoref{fig:pmf-cmp}, we show the shapes of
the COM-Poisson$_\mu$ distribution.

@@

The constraint $\mu>0$ imposes an undesirable restriction in the
original parameter space, $\lambda^\nu>(\nu-1)/(2\nu)$. However, this
restricted region is related to small underdispersed counts (averages
smaller than 0.1 and $\nu>1$) a parameter region unlikely to be of
interest in practice.

In order to explore the flexibility of the COM-Poisson model to deal
with real count data, we compute indexes for dispersion (DI),
zero-inflation (ZI) and heavy-tail (HI), which are respectively given by
\begin{equation*}
\text{DI} = \frac{\text{Var}(Y)}{\text{E}(Y)}, \quad
\text{ZI} = 1 + \frac{\log \Pr(Y = 0)}{\text{E}(Y)}
  \quad \text{and} \quad
\text{HT} = \frac{\Pr(Y=y+1)}{\Pr(Y=y)}\quad \text{for} \quad y \to
\infty.
\end{equation*}
These indexes are defined in relation to the Poisson distribution. Thus,
the dispersion index indicates overdispersion for $\text{DI} > 1$,
underdispersion for $\text{DI} < 1$ and equidispersion for
$\text{DI} = 1$. The zero-inflation index indicates zero-inflation for
$\text{ZI} > 0$, zero-deflation for $\text{ZI} < 0$ and no excess of
zeros for $\text{ZI} = 0$. Finally, the heavy-tail index indicates a
heavy-tail distribution for $\text{HT} \to 1$ when $y \to \infty$.
These indexes are discussed by \citet{Bonat2017} to study the
flexibility of Poisson-Tweedie distribution, and \citet{Puig2006} to
describe count distributions in general.

@@

<<indexes-plot, fig.width=9.5, fig.height=4, fig.cap="Indexes for the COM-Poisson distribution. (a) Mean and variance relationship, (b--d) dispersion, zero-inflation and heavy-tail indexes for different parameter values. Dotted lines represents the special case of the Poisson distribution.">>=
@
